---
title: "R Notebook"
output: html_notebook
---


Load data
```{r}
#import libraries and data 
library(psych)
library(dplyr)
library(mirt)
library(readxl)
#load data 
full_response <- read_csv("CBC_CAT_responses_full.csv")
aptitude <- read_csv("CBC_CAT_aptitude_responses.csv")
personality <- read_csv("CBC_CAT_personality_responses.csv")
mapping_table <- read_excel("mapping_table.xlsx")

#preview first few rows
head(full_response)
head(aptitude)
head(personality)
head(mapping_table)

```
# Data Quality Check. 

## a) Reliability: Crobach's Alpha

```{r}
#ignore student id before running alphas
full_items <- full_response %>%
  select(-student_id) %>%
  mutate(across(everything(), as.numeric))

aptitude_items <- aptitude %>%
  select(-student_id) %>%
  mutate(across(everything(), as.numeric))

personality_items <- personality %>%
  select(-student_id) %>%
  mutate(across(everything(), as.numeric))


#run cronbach's alpha
alpha(full_items, check.keys = TRUE)
alpha(aptitude_items, check.keys = TRUE)
alpha(personality_items, check.keys = TRUE)

#item correlation check 
alpha_full <- alpha(full_items, check.keys = TRUE)
alpha_apt <- alpha(aptitude_items, check.keys = TRUE)
alpha_pers <- alpha(personality_items, check.keys = TRUE)

#results
alpha_full$item.stats
alpha_apt$item.stats
alpha_pers$item.stats

```

Before assessing reliability, non-item identifier variables were removed and all item responses were converted to numeric format. Reverse-keyed items were handled using check.keys = TRUE. Internal consistency was then assessed using Cronbach’s alpha for the full item pool, aptitude subscale, and personality subscale.

Reliability of the generated response data was assessed using Cronbach’s alpha for the full item pool and each subscale.

For the full dataset (56 items), the standardized Cronbach’s alpha was 0.83, indicating good internal consistency after accounting for differences in item variance.

The aptitude subscale (30 items) demonstrated excellent reliability (α = 0.90), suggesting a high level of consistency among items measuring cognitive ability.

The personality subscale (26 items) showed moderate reliability (α = 0.61). This is expected, as personality constructs are typically more heterogeneous and may capture multiple underlying traits. It may also reflect the presence of reverse-coded or multidimensional items, which is common in personality measurement.

Overall, the results indicate that the generated data exhibit acceptable to strong reliability, supporting their suitability for subsequent IRT calibration.

b) Validity

Construct Validity
```{r}
#validity of the full dataset
fa.parallel(full_items)
fa_full <- fa(full_items, nfactors = 2, rotate = "varimax")
print(fa_full$loadings, cutoff = 0.3)
```
Factor analysis of the full dataset revealed a two-factor structure corresponding to the aptitude and personality item groups, supporting the intended construct design.

```{r}
#subscale
#aptitutde 
fa(aptitude_items, nfactors = 1)

#personality 
fa(personality_items, n_factors = 1)
```
Factor analysis of the full dataset revealed a two-factor structure corresponding to the aptitude and personality item groups, supporting the intended construct design.


Discriminant validity 

```{r}
#discriminant validity 
apt_score <- rowSums(aptitude_items)
pers_score <- rowSums(personality_items)

cor(apt_score, pers_score)
```
The correlation between aptitude and personality scores was very low (r = 0.03), indicating that the two constructs are largely independent. This supports strong discriminant validity, as the two item sets are designed to measure distinct underlying traits.


Content validty 
```{r}
#aptitutde difficulty 
colMeans(aptitude_items)
#personality varibaility 
apply(personality_items, 2, sd)
```
The proportion of correct responses for aptitude items ranged from approximately 0.10 to 0.80, indicating a well-distributed range of item difficulty. Easier items had higher response rates, while more difficult items had lower response rates, consistent with the simulated design.

Personality items exhibited consistent variability, with standard deviations ranging from approximately 1.18 to 1.39. This indicates that responses are well-distributed across respondents and that all items contribute meaningful information.

## Calibrate Item Parameters

```{r}
#prep aptitutde data 
apt_matrix <- as.matrix(aptitude_items)
#fit irt model 
model_2pl <- mirt(apt_matrix, 1, itemtype = "2PL")
#extract item parameters
coef_2pl <- coef(model_2pl, IRTpars = TRUE, simplify = TRUE)
coef_2pl$items

#estimate student ability 
theta <- fscores(model_2pl)
```

The 2PL model results indicate that item discrimination parameters range from approximately 0.76 to 2.24, suggesting that most items have good to strong ability to differentiate between students. Item difficulty parameters span from -2.09 to 1.87, covering a wide range from easy to difficult items. This distribution is appropriate for computerized adaptive testing, as it allows accurate measurement across different ability levels.


Curves 
```{r}
#ICC 
plot(model_2pl, type = "trace")

#IIC
plot(model_2pl, type = "info")
```

## Build adaptive testing system 
```{r}
#create function 
run_cat <- function(model, response_vector, max_items = 10, se_stop = 0.30) {
  
  all_items <- colnames(response_vector)
  administered <- c()
  responses <- c()
  theta_history <- c()
  se_history <- c()
  
  theta_est <- 0
  se_est <- NA
  
  params <- coef(model, IRTpars = TRUE, simplify = TRUE)$items
  
  for (i in 1:max_items) {
    
    # choose first item: medium difficulty
    if (i == 1) {
      next_item <- rownames(params)[which.min(abs(params[, "b"]))]
    } else {
      unused_items <- setdiff(all_items, administered)
      
      info_values <- sapply(unused_items, function(item) {
        item_index <- which(all_items == item)
        iteminfo(extract.item(model, item_index), Theta = theta_est)
      })
      
      next_item <- unused_items[which.max(info_values)]
    }
    
    # administer item
    administered <- c(administered, next_item)
    responses <- c(responses, as.numeric(response_vector[[next_item]]))
    
    # create response pattern with NA for unadministered items
    pattern <- rep(NA, length(all_items))
    names(pattern) <- all_items
    pattern[administered] <- responses
    
    # update theta
    score <- fscores(
      model,
      response.pattern = matrix(pattern, nrow = 1),
      method = "EAP",
      full.scores.SE = TRUE)
    
    theta_est <- score[1, 1]
    se_est <- score[1, 2]
    
    theta_history <- c(theta_history, theta_est)
    se_history <- c(se_history, se_est)
    
    if (!is.na(se_est) && se_est < se_stop) break
  }
  
  data.frame(
    step = 1:length(administered),
    item = administered,
    response = responses,
    theta = theta_history,
    SE = se_history
  )
}

#run cat for one student 
student1 <- aptitude_items[1, , drop = FALSE]

#run function 
cat_result <- run_cat(model_2pl, student1, max_items = 10, se_stop = 0.30)

#results 
cat_result
```
The CAT system dynamically updates the student’s ability estimate after each item. As more items are administered, the standard error decreases, indicating increased precision. The sequence of items adapts to the student’s responses, selecting items that maximize information at the current ability estimate.


Run for 3 students 
```{r}
#three example students
#compute total score for each student
scores <- rowSums(aptitude_items)
#identify low, middle, and high student abaility 
low_id <- which.min(scores)
high_id <- which.max(scores)
mid_id <- which.min(abs(scores - mean(scores)))
#run Cat for each selected student 
cat_low  <- run_cat(model_2pl, aptitude_items[low_id, , drop = FALSE])
cat_mid  <- run_cat(model_2pl, aptitude_items[mid_id, , drop = FALSE])
cat_high <- run_cat(model_2pl, aptitude_items[high_id, , drop = FALSE])

#show results 
print(cat_low)
print(cat_mid)
print(cat_high)
```

```{r}
#compare final ability 
cat_low$theta[nrow(cat_low)]
cat_mid$theta[nrow(cat_mid)]
cat_high$theta[nrow(cat_high)]

#summary table
data.frame(
  student = c("Low", "Mid", "High"),
  final_theta = c(
    tail(cat_low$theta, 1),
    tail(cat_mid$theta, 1),
    tail(cat_high$theta, 1)
  ),
  final_SE = c(
    tail(cat_low$SE, 1),
    tail(cat_mid$SE, 1),
    tail(cat_high$SE, 1)
  )
)
```
Pathway recommendation 
```{r}
#pathway function
recommend_pathway <- function(cat_result, mapping_table) {
  
  # merge CAT results with mapping table
  df <- merge(cat_result, mapping_table,
              by.x = "item", by.y = "Item ID")
  
  # initialize scores
  STEM <- 0
  SOCIAL <- 0
  ARTS <- 0
  
  # loop through items
  for (i in 1:nrow(df)) {
    
    # weight by response (0 or 1)
    if (df$`Pathway Signal`[i] == "STEM") {
      STEM <- STEM + df$response[i]
    }
    if (df$`Pathway Signal`[i] == "Social Sciences") {
      SOCIAL <- SOCIAL + df$response[i]
    }
    if (df$`Pathway Signal`[i] == "Arts & Sports") {
      ARTS <- ARTS + df$response[i]
    }
  }
  
  # normalize by number of items asked per pathway
  STEM <- STEM / sum(df$`Pathway Signal` == "STEM")
  SOCIAL <- SOCIAL / sum(df$`Pathway Signal` == "Social Sciences")
  ARTS <- ARTS / sum(df$`Pathway Signal` == "Arts & Sports")
  
  # combine scores
  scores <- c(STEM = STEM, SOCIAL = SOCIAL, ARTS = ARTS)
  
  # recommendation
  if (all(scores == 0)) {
    recommendation <- "No clear recommendation"
  } else {
    recommendation <- names(scores)[which.max(scores)]
  }
}

#apply the function 
rec_low  <- recommend_pathway(cat_low, mapping_table)
rec_mid  <- recommend_pathway(cat_mid, mapping_table)
rec_high <- recommend_pathway(cat_high, mapping_table)

#output
print(rec_low)
print(rec_mid)
print(rec_high)

```

```{r}

```